---
title: "The factors of students’ performance"
author: 
 - Stanley Njiru Kariuki
 - Ahmed Hamouda 
 - Xiaowei Wan
 - Yulin Chen
date: "6/18/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align="center")
```

## Abstract
This study seeks to determine the possible correlation among several factors affecting student academic performance. Data will be collected from a learning management system (LMS) called Kallboard 360, which provides users with a synchronous access to educational resources. The dataset consists of 480 student records of males and females. It also contains demographic, academic, and behavioral information from 14 different countries. R programming software will be utilized to conduct analysis and measure variables associated with academic performance. We hypothesize that the more active a student is in the class, the better performance they will have. In addition, this study will observe parent cooperation with the school relative to better performance outcomes.


## Introduction
Academic performance has been a standard measure for many countries and academic systems throughout the world. It is used to determine entry into higher institutions, strengths and weaknesses of students, along with directing job placements within certain fields.  In this case, academic performance is the extent to which students, have achieved their educational goals. Grading has long been used as a method to rank student achievement and performance.


However, it should be noted that there are several factors which have an effect on performance standards ranging from environment, economic resources, educational systems, along with societal beliefs and expectations. For example, a study conducted by Farooq, Chaudhry, Shafiq and Berhanu (2011) looked at factors affecting students’ quality of academic performance, and found that socio-economic status and parents’ perception of education had a significant effect on overall academic achievement in secondary level education. This makes it challenging to specify direct factors which contribute to this. The purpose of this study is to determine attributes affecting student’s academic performance at lower level, middle school and high school. This will allow for us to determine which factors are the most indicative of academic performance. 
Thus, the hypothesis that the more active a student is in the class, the better performance they will have and that the more a parent cooperates with the school the better the student’s academic performance. 

### Read dataset and summary of edu dataset
```{r student performance dataset}
# Read dataset to edudataset
studataset <- read.csv(file="/Users/xwan/Documents/HU/ANLT 510-51/xAPI-Edu-Data.csv",head=TRUE,sep=",")
# To get definition of the columns type
str(studataset)
```

### Summary of Edu Dataset
```{r Edu dataset summary}
# Summary of Edu Dataset
summary(studataset)
```
### Catagorical variables
```{r edu dataset catagorical plots}
par(mfrow = c(2, 2))
plot(studataset$gender, sub="Gender", col=2:6)
plot(studataset$NationalITy, sub="NationalITy", col=2:6)
plot(studataset$PlaceofBirth, sub="PlaceofBirth", col=2:6)
plot(studataset$StageID, sub="StageID", col=2:6)
plot(studataset$Relation, sub="Relation", col=2:6)
plot(studataset$GradeID, sub="GradeID", col=2:6)
plot(studataset$Topic, sub="Topic", col=2:6)
plot(studataset$Semester, sub="Semester", col=2:6)
```
```{r student absence days}
par(mfrow = c(1, 2))
plot(studataset$StudentAbsenceDays, sub="StudentAbsenceDays", col=2:6)
```

### Based on the the above plots several observations can be made:


* Majority of students are male
* Most of the students are from either Kuwait or Jordan
* Majority of students are middleschool or lower

### Numerical Varaibles
```{r numerical varaibles}
par(mfrow = c(2,2))
plot(studataset$Class ~ studataset$raisedhands, xlab="raisedhands", ylab = "Class", col=2:6)
plot(studataset$Class ~ studataset$VisITedResources, xlab="VisITedResources", ylab = "Class", col=2:6)
plot(studataset$Class ~ studataset$AnnouncementsView, xlab="AnnouncementsView", ylab = "Class", col=2:6)
plot(studataset$Class ~ studataset$Discussion, xlab="Discussion", ylab = "Class", col=2:6)
```

In general these features represent the level of student particapation in their respective courses. Cleary the students with low grades tend to have lower level of class particapation.

## Methods
The data was collected from learning management system (LMS) called Kallboard 360. Such system provides users with a synchronous access to educational resources from any device with internet connection. The dataset consists of 480 student records and 16 features. The features are divided into three major categories:


1. Demographic features such as gender and nationality.
2. Academic background features such as educational stage, grade Level and section. 
3. Behavioral features such as raised hand on class, opening resources, and school satisfaction.


The dataset consists of 305 males and 175 females. The students come from different origins such as 179 students are from Kuwait, 172 students are from Jordan, 28 students from Palestine, 22 students are from Iraq, 17 students from Lebanon, 12 students from Tunis, 11 students from Saudi Arabia, 9 students from Egypt, 7 students from Syria, 6 students from USA, Iran and Libya, 4 students from Morocco and one student from Venezuela.


The dataset is collected through two educational semesters: 245 student records are collected during the first semester and 235 student records are collected during the second semester. The data set includes also the school attendance feature such as the students are classified into two categories based on their absence days: 191 students exceed 7 absence days and 289 students their absence days under 7.


## Discussion
The above study is conducted to observe the factors that can influence student performance. The performance of the student is evaluated as class score, low, middle, or high. In our research, we analyze different factors among students who work good, in the middle, and who have a lower score.


We set up a couple of different categories of factors to conduct the analysis, external, internal and parents’ performance. We know every kid has different talent, from the research we did, we found that generally, girls perform better than boys. It is interesting that, contrary to popular perception, girls perform much better than boys in science class. We believe as our research goes on, we will discover more phenomenon like this, which will make our research more and more interesting.


Meanwhile, we understand that performance of the students in academics is not only influenced by their own characteristics gifted by nature but also effortful behavior in class is involved, for example, the frequency of raising hands, participating in the discussion, student absence days, and etc. Furthermore, how a parent performs will also impact a student.  We also take the parents performance into consideration.


For the next step, we will focus our study on how active and energetic a student performs in class.  We make the following hypothesis:


  * The more active a student is in the class, the better outcome of the performance they will gain.
  * The more the parent cooperate with the school, the better performance the student will have.
  
  
At the same time, we will focus on the study of the correlations between the various factors. And we will also analyze that among all these different factors, which one has the biggest impact to students’ performance.


Hopefully, on the basis of our study and conclusions, recommendations can be suggested in order to solve the problems and limitations that are responsible for low student’s performance. We believe these recommendations will be useful for both administrative and academic point of view.


## References


Student's Academic Performance Dataset (xAPI-Edu-Data)


https://www.kaggle.com/aljarah/xAPI-Edu-Data


Daniyal, M. (n.d). The factors affecting the students’ performance. Retrieved from academia:


https://www.academia.edu/4729029/THE_FACTORS_AFFECTING_THE_STUDENTS_PERFORMANCE


Rmarkdown. (n.d.). HTML Documents. Retrieved from Rmarkdown Rstudio:


http://rmarkdown.rstudio.com/html_document_format.html#custom_css


M.S. Farooq, A.H. Chaudhry, M. Shafiq, G. Berhanu Factors Affecting Students’ Quality Of Academic Performance: A Case Of Secondary School Level 


http://pu.edu.pk/images/journal/iqtm/PDF-FILES/01-Factor.pdf
